---
title: "Nonresponse Adjustments"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
vignette: >
  %\VignetteIndexEntry{Nonresponse adjustments}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=FALSE, warning=FALSE}
library(dplyr) # For data manipulation
library(survey) # For complex survey analysis
library(srvyr) # For complex survey analysis with dplyr syntax
library(svrep)
```

In this short vignette, we'll demonstrate how the svrep package can be used to implement weighting adjustments for nonresponse in survey samples. For illustration, we'll use an example survey measuring Covid-19 vaccination status and a handful of demographic variables, based on a simple random sample of 1,000 residents of Louisville, Kentucky.

```{r}
# Load and inspect the data
data("lou_vax_survey", package = 'svrep')
head(lou_vax_survey)
colnames(lou_vax_survey)
```

This vaccination survey has an overall response rate of 50.2%, which means that estimated vaccination rates may be significantly biased by nonresponse. We'll use nonresponse weighting adjustments to try and reduce potential nonresponse bias.

```{r}
lou_vax_survey |> count(RESPONSE_STATUS) |> mutate(pct = n/sum(n))
```

## Creating initial replicate weights

To begin with, we'll create bootstrap replicate weights using the `as.svrepdesign()` function from the survey package. The initial step is to describe the survey design using the `svydesign()` function and then create appropriate replicate weights using `as.svrepdesign()` and additional arguments such as `type` (with options `'JK1'`, `'JKn'`, `'bootstrap'`, `'BRR'`, `'Fay'`, etc.) At the beginning, the replicate weights allow us to estimate variance in unadjusted estimates caused by random sampling. Note that when creating the initial replicate weights, it is almost always necessary to use data for the entire selected sample (i.e. respondents as well as nonrespondents).

```{r}
# Describe the survey design
lou_vax_survey <- svydesign(ids = ~ 1, weights = ~ SAMPLING_WEIGHT,
                            data = lou_vax_survey)

print(lou_vax_survey)

# Create appropriate replicate weights
lou_vax_survey <- lou_vax_survey |>
  as.svrepdesign(type = "boot", replicates = 100, mse = TRUE)

print(lou_vax_survey)
```

For convenience, we'll convert this survey design object into an object with class `tbl_svy`, which allows us to use convenient tidyverse/dplyr syntax (`group_by()`, `summarize()`, etc.) as well as other helpful functions from [the srvyr package](https://cran.r-project.org/web/packages/srvyr/vignettes/srvyr-vs-survey.html).

```{r}
lou_vax_survey <- lou_vax_survey |> as_survey()

print(lou_vax_survey)
```

## Redistributing weight from nonrespondents to respondents

A common form of nonresponse adjustment is to simply 'redistribute' weight from the nonrespondents to the nonrespondents. In other words, the weight for each nonrespondent is set to $0$, and the weight for each respondent is increased by a factor greater than one so that the sum of adjusted weights in the sample of respondents equals the sum of unadjusted weights from the full sample. For example, if the sum of weights among respondents is $299,544.4$ and the sum of weights among nonrespondents is $297,157.6$, then a basic nonresponse adjustment would set the weights among nonrespondents to $0$ and multiply the weight for each respondent by an adjustment factor equal to $1 + (297,157.6/299,544.4)$. This type of adjustment is succinctly described in mathematical notation below.

```{=tex}
\begin{aligned}
w_i &= \textit{Original sampling weight for case }i \\
    &= 1/\pi_i, \textit{ where } \pi_i \textit{ is the probability case i }\textit{had of being sampled}\\
f_{NR,i} &= \textit{Nonresponse adjustment factor for case i} \\
w_{NR, i} &= w_i \times f_{NR,i} = \textit{Weight for case }i \textit{ after nonresponse adjustment} \\

\\
\sum_{i \in s_{resp}} w_i &= \textit{Sum of sampling weights among respondents} \\
\sum_{i \in s_{nonresp}} w_i &= \textit{Sum of sampling weights among nonrespondents} \\
\\
f_{NR_i} &= \begin{cases} 
     0 & \textit{if case }i\textit{ is a nonrespondent} \\
     1 + \frac{\sum_{i \in s_{nonresp}} w_i}{\sum_{i \in s_{resp}}} & \textit{if case }i\textit{ is a respondent} 
           \end{cases}
\end{aligned}
```

We'll illustrate this type of adjustment with the Louisville vaccination survey. First, we'll inspect the sum of the sampling weights for respondents, nonrespondents, and the overall sample.

```{r}
# Weights before adjustment
lou_vax_survey |>
  group_by(RESPONSE_STATUS) |>
  cascade(
    `Sum of Weights` = sum(cur_svy_wts()),
    .fill = "TOTAL"
  )
```

Next, we'll redistribute weight from nonrespondents to respondents using the `redistribute_weights()` function, which adjusts the full-sample weights as well as each set of replicate weights. To specify which subset of data should have its weights reduced, we supply a logical expression to the argument `reduce_if`. To specify which subset of data should have its weights increased, we supply a logical expression to the argument `increase_if`.

```{r}
# Conduct a basic nonresponse adjustment
nr_adjusted_survey <- lou_vax_survey |>
  redistribute_weights(
    reduce_if = RESPONSE_STATUS == "Nonrespondent",
    increase_if = RESPONSE_STATUS == "Respondent"
  )
```

After making the adjustment, we can check that all of the weight from nonrespondents has been redistributed to respondents.

```{r}
# Check the sum of full-sample weights by response status
nr_adjusted_survey |>
  group_by(RESPONSE_STATUS) |>
  cascade(
    `Sum of Weights` = sum(cur_svy_wts()),
    .fill = "TOTAL"
  )
```

```{r}
# For each replicate, get sum of weights by response status
replicate_weight_sums_by_response_status <- svytotal(
  x = ~ RESPONSE_STATUS,
  design = nr_adjusted_survey,
  return.replicates = TRUE
) |> 
  getElement('replicates') |> # Extract totals from each replicate
  `rownames<-`(paste0("Replicate ", 1:100)) |> # Add helpful row names
  `colnames<-`(c("Nonrespondent", "Respondent")) # Add helpful column names

# Preview first few replicates
writeLines("Sum of weights by response status and bootstrap replicate")
replicate_weight_sums_by_response_status |> head()
```

The motivation for making this adjustment is that standard methods of statistical inference assume that every person in the population has a known, nonzero probability of participating in the survey (i.e. has a nonzero chance of being sampled and a nonzero chance of responding if they are sampled), denoted $p_{i,overall}$. Basic results in survey sampling theory guarantee that if this assumption is true, we can produce unbiased estimates of population means and totals by weighting data from each respondent with the weight $1/{p_{i,overall}}$.

that this probability can be used to produce unbiased estimates of population totals by weighting data from . In order to use these standard methods of inference then, we must estimate an overall probability of participating in the survey.

 For respondent $i$, this probability is defined as $\pi_i \times p_i$, their probability of being randomly sampled multiplied by their probability of responding to the survey if sampled. A simple estimate of $p_i$ is the weighted response rate $\frac{\sum_{i \in s_{resp}}w_i}{\sum_{i \in s}w_i}$. 


When we use the nonresponse-adjusted weights for analyses, we need to take the adjustment into account during variance estimation. Because of random sampling, the precise adjustment factor used in the nonresponse adjustment will vary from one sample to the next. To account for this variability, we can apply the nonresponse adjustment not only to the full-sample weights, but to each set of replicate weights as well. The `redistribute_weights()` function handles this for us: after nonresponse adjustment, all of the weight in each replicate has been redistributed from nonrespondents to respondents.


## Conducting weighting class adjustments

Nonresponse bias is liable to occur if different subpopulations systematically differ in terms of their response rates to the survey and also differ in terms of what the survey is trying to measure (in this case, vaccination status). In our example, we can see some fairly large differences in response rates across different race/ethnicity groups.

```{r}
lou_vax_survey |>
  group_by(RACE_ETHNICITY) |>
  summarize(Response_Rate = mean(RESPONSE_STATUS == "Respondent"),
            Sample_Size = n(),
            n_Respondents = sum(RESPONSE_STATUS == "Respondent"))
```

Weighting adjustments may be able to help reduce nonresponse bias caused by these differences in response rates. One standard form of adjustment known as **weighting class adjustment** is to redistribute weights from nonrespondents to respondents separately by different categories of auxiliary variables (such as race/ethnicity). Chapter 2 of Heeringa, S., West, B., Berglund, P. (2017) provides an excellent overview of weighting class adjustments. To implement a weighting class adjustment with the svrep package, we can simply use the `by` argument of `redistribute_weights()`.

```{r}
nr_adjusted_survey <- lou_vax_survey |>
  redistribute_weights(
    reduce_if = RESPONSE_STATUS == "Nonrespondent",
    increase_if = RESPONSE_STATUS == "Respondent",
    by = c("RACE_ETHNICITY")
  )
```

Multiple grouping variables may be supplied to the `by` argument. For example, one can specify `by = c("STRATUM", "RACE_ETHNICITY")` to redistribute weights separately by combinations of stratum and race/ethnicity category.

## Saving the final weights to a data file

Once we're satisfied with the weights, we can create a data frame with the analysis variables and columns of replicate weights. This format is easy to export to data files that can be loaded into R or other software later.

```{r}
data_frame_with_nr_adjusted_weights <- nr_adjusted_survey |>
  as_data_frame_with_weights(
    full_wgt_name = "NR_ADJ_WGT",
    rep_wgt_prefix = "NR_ADJ_REP_WGT_"
  )

# Preview first 10 column names
colnames(data_frame_with_nr_adjusted_weights) |> head(10)
```

```{r, eval=FALSE}
# Write the data to a CSV file
write.csv(
  x = data_frame_with_nr_adjusted_weights,
  file = "survey-data_with-nonresponse-adjusted-weights.csv"
)
```

## Recommended Reading

-   See Chapter 2, Section 2.7.3 of "Applied Survey Data Analysis" for a statistical explanation of the weighting adjustments described in this vignette.

> Heeringa, S., West, B., Berglund, P. (2017). Applied Survey Data Analysis, 2nd edition. Boca Raton, FL: CRC Press.

-   Chapter 13 of "Practical Tools for Designing and Weighting Survey Samples" also provides an excellent overview of the nonresponse adjustment methods.

> Valliant, R., Dever, J., Kreuter, F. (2018). Practical Tools for Designing and Weighting Survey Samples, 2nd edition. New York: Springer.

-   Rust K, Rao J. (1996). Variance estimation for complex surveys using replication techniques. Statistical Methods in Medical Research. 5(3):283-310. <doi:10.1177/096228029600500305>.
